Control Aware Radio Resource Allocation in Low Latency Wireless Control Systems

### 2019-02-05 Low_latency_control/response/response_main.tex

\issue{ It was unclear whether the focus is on communication for a given control or the other way around. Taht is currently it looks like the solution is based on sequential design.}
\answer{We thank the reviewer for brining up this point of clarification. The focus of this work is designing a scheduling protocol for a given system and control. As such, the control is assumed to be given, or chosen agnostic to the communication, while the communication is designed relative to both the system dynamics and choice of control. We have added a sentence to Section II to clarify this point. }

\issue{ The article would improve a lot if specific use cases are defined.}
\answer{\red{To be written}}

\issue{ Linear control was adopted (for tractability) but clearly the nonlinearity is needed}
\answer{The reviewer brings up a reasonable point about the formulation of the wireless control problem. Indeed, many practical systems feature both nonlinear dynamics and potentially non-linear controls. As stated, the proposed approach focuses on linear systems and controls so that we may derive analytic expressions for the control-aware adaptive packet delivery rates. In practice, nonlinear systems may be linearized for the purposes of obtaining such packet delivery rate targets in (23) and performing the proposed CALLS method as written. \blue{To demonstrate how this may be done, in the revised manuscript we have included additional simulations in Section VI in which we control a non-linear system using the CALLS method to schedule plant communications. In these simulations, while the actual dynamics and control being applied are nonlinear in nature, a linearization of the dynamics is used solely to derive the necessary PDR targets as in expression (23). Here, we demonstrate that the proposed method is applicable even in nonlinear systems.}   }

\issue{Reliability as not addressed however it is intertwined with latency}
\answer{This is a good point raised by the reviewer that deserves clarification. Reliability is a key component of a practical system that must be carefully managed in low latency systems. In our work, rather than aim for traditional, communication-based reliability targets, such as $0.999$ packet delivery rates, we consider the fact that the observed performance of a control system is more closely related to a \emph{control}-based reliability metric, such as stability or distance from origin. \blue{As discussed in Section...}, by defining reliability targets with respect to the Lyapunov cost function rather than with respect to a fixed packet delivery rate target, we are able to derive \emph{adaptive} packet delivery rate requirements that change with the current performance of each control system. When a control system is in a safe or stable range, it can remain in good performance with smaller packet delivery rates than a control system in a less desirable region. Thus, the proposed CALLS method aims to achieve reliability of control system performance, in other words an \emph{adaptive reliability}, in contrast to a standard high reliability framework. Such adaptive rates are critical in enabling low-latency transmissions in large-scale wireless settings, as observed our simulations in Section \blue{???}. We point out that the proposed control-adaptive framework is thus presented as an alternative to high reliability scheduling schemes in settings in which the control systems are well-managed and defined, such as in the case of industrial controls.  \blue{We have added a discussion of this to the revised manuscript.}}

### 2019-02-12 Low_latency_control/revised/conclusions.tex
Add files via upload

The results presented in this paper suggest an interesting
potential for control-aware resource allocation and scheduling, particularly in low-latency industrial systems. By considering the control-specific targets such as maintaining
stability or an error margin, we observe that the standard
high reliability targets considered (e.g. packet delivery rates $\geq 0.999$) can in some cases be substantially stricter than
necessary for adequate performance. Wireless control systems
with sufficiently slow dynamics can be kept stable with much
lower packet delivery rates, which in turn make low-latency
communications more achievable. Furthermore, in realistic industrial systems there will be many heterogeneous devices being controlled, whose variation in communication needs is well-served by control-aware opportunism proposed in this paper. This suggests the potential for wireless communications to be adopted using a smart control-communication co-design approach even
while ultra-reliable wireless system technology remains under development.

>>>>

\red{The results presented in this paper suggest an interesting
potential for control-aware resource allocation and scheduling, particularly in low-latency industrial systems. By considering the control-specific targets such as maintaining
stability or an error margin, we observe that the standard
high reliability targets considered in URLCC (e.g. packet delivery rates $\geq 0.999$) can in some cases be substantially stricter than
necessary for adequate performance. Wireless control systems
with sufficiently slow dynamics can be kept stable with much
lower packet delivery rates, which in turn make low-latency
communications more achievable.} Furthermore, in realistic industrial systems there will be many heterogeneous devices being controlled, whose variation in communication needs is well-served by control-aware adaptivity proposed in this paper. This suggests the potential for wireless communications to be adopted using a smart control-communication co-design approach even
while ultra-reliable wireless system technology remains under development.

### 2019-02-12 Low_latency_control/revised/introduction.tex
Add files via upload

The reliability and robustness of wireless communication systems has become a key component in the design and implementation of large scale systems in the Internet-of-Things (IoT). For these systems to operate successfully, it is important that the data collected by sensors can be communicated throughout the IoT network. While the most reliable form of data communication is through wired connections, the scale and mobility of modern IoT settings has rendered the cost of installing and maintaining a wired network a significant challenge~\cite{wollschlaeger2017future}. Consequently, there is great interest in the design of wireless control systems that can achieve reliable performance~\cite{varghese2014wireless, li2017review}. Low-latency wireless transmissions are however a necessary feature in many wireless IoT systems, particularly those in industrial control~\cite{varghese2014wireless,popovski2018wireless, bennis2018ultra}.  The primary challenge is then the inherent trade-off that occurs between reliability and latency, as it is difficult to maintain high reliability while using higher data rates due to the stochasticity of the wireless channel. It is necessary to design resource allocation and scheduling protocols for wireless control systems that can both meet reliability \emph{and} latency requirements of the control system.

In the wireless communications research and industry, many radio resource allocation schemes in the form of wireless scheduling techniques have been proposed to provide reliability, or quality of service (QoS), to users across the network in the form of throughput, fairness and/or latency~\cite{cao2001scheduling,wongthavarawat2003packet,fattah2002overview,yaacoub2012survey}. For instance, round-robin scheduling is a common approach due to its simplicity inherent fairness. Algorithms such as proportional fair (PF)~\cite{kim2005proportional}, max-SNR~\cite{knopp1995information} and their variations have been designed for maximizing the overall system throughput and/or fairness. For time-sensitive applications such as industrial control and automation, the experiences of individual users/devices are highly dependent on the ability of the network to deliver their packets within a latency bound. Generic delay-aware schedulers such as EDF~\cite{wu2014analysis} and WFQ~\cite{lu1999fair} do not adapt to wireless channel conditions, while channel state information is considered in M-LWDF \cite{andrews2001providing}. However, all such methods are unable to leverage the information on the control system dynamics and thus can make unsuitable radio resource allocation decisions when deployed in time-sensitive wireless control systems.

In the context of wireless control systems, there have been a range of works that incorporates control system information in the networking and communication policies.
%
The mechanisms usually examined are either static or dynamic. Typical examples of the former type are periodically protocols where the wireless devices transmit in a predefined repeating order, e.g., round-robin. Control system stability under such protocols can be analyzed -- see, e.g.,~\cite{Hespanha_survey, Schenato_foundations, Donkers_switched, Branicky_stability}. Periodic sequences leading to stability~\cite{Hristu_shared_feedback}, controllability and observability~\cite{Hristu_communication_control}, or optimizing control objectives~\cite{LeNy_resource_LQR,Meier_measurement_control, Scheduling_control_combinatorics} have been proposed.
%
Dynamic schedulers do not rely on a predefined sequence but decide access to the communication medium dynamically at each step. %, for example by dynamically assigning priorities to the competing tasks.
Initial approaches abstract control performance requirements in the time/frequency domain, e.g., how often a task needs resource access, employing algorithms from real-time scheduling theory~\cite{Branicky_RM, Liu_Real_time_systems}. More recent scheduling approaches often depend on the current control system states, i.e., informally the subsystem with the largest state discrepancy is scheduled to communicate -- see, e.g.,~\cite{Donkers_switched, Walsh_stability, Hristu_Kumar_interrupt_based, Egerstedt_queue, Hirche_Scheduling_Price, Cervin_event_scheduling, mamduhi2014event,shi2011optimal,han2017optimal}.
%
Alternatively scheduling can take into account current wireless channel conditions opportunistically to meet target control system reliability requirements~\cite{GatsisEtal15}. None of these approaches, however, are explicitly designed to achieve strong performance it low-latency scenarios.

In this paper, we develop a control-aware, high reliability, low-latency IEEE 802.11ax WiFi protocol~\cite{bellalta2016ieee} that is designed to reduce the total transmission time at each uplink cycle in the control loop. This is done through the mathematical formulation of the control system design goal in the form of a Lyapunov function that ensures stability of the control system. This formulation naturally induces a bound on the packet delivery rate each control system needs to achieve to meet the control-based goal. Such packet delivery rates depend upon current control and channel states and thus dynamically change over the course of the system life time. This can be viewed as an opportunistic protocol with respect to both the current control states \emph{and} channel conditions. Furthermore, because these control-based success rate requirements may be significantly lower than traditional, high reliability communication demands, the proposed method is better suited to find scheduling configurations that can moreover meet the strict latency requirements imposed by the physical system. This is in contrast to the control-aware approach taken in \cite{GatsisEtal15}, which focuses on low-power and infrequent transmissions rather than the low-latency setting of interest in industrial control.

@@

Using this formulation, we develop the control-aware low latency scheduling (CALLS) method (Section \ref{sec_calls}). The CALLS method uses current control states and channel conditions to derive dynamic packet success rates for each user (Section \ref{sec_psr}). In this way, control systems that are closest to instability will be given priority in the scheduling so that they may close their control loops. The scheduling procedure consists of a random user selection procedure to reduce the number of required PPDUs that incur significant overhead  (Section \ref{sec_rss}), followed by an assignment-method based scheduling of selected users to minimize total transmission time (Section \ref{sec_assignment}). The performance of the CALLS method is analyzed in a series of simulation experiments in which its performance is compared against a control-agnostic procedure (Section \ref{sec_simulation}). We demonstrate in numerous low-latency control systems that the control-aware method can support more users than the alternative and achieve more robust overall performance.

%The reliability and robustness of wireless communication systems has become a key component in the design and implementation of large scale systems in the Internet-of-Things (IoT). For these systems to operate successfully, it is important that the data or information collected by sensors can be communicated throughout the IoT network. While the most reliable form of data communication is through wired connections, the scale and mobility of modern IoT settings has rendered the cost of installing and maintaining a wired network a significant challenge~\cite{wollschlaeger2017future}. Consequently, the design of wireless control systems that can achieve reliable performance has become of great interest~\cite{varghese2014wireless, li2017review}. The primary challenge is then the inherent trade off that occurs between reliability and latency. Low-latency transmissions are a necessary feature in many wireless IoT systems, particularly those in industrial control~\cite{varghese2014wireless}. It is difficult to maintain high reliability while using higher data rates, however, due to the stochasticity of the wireless channel. It is necessary then to design resource allocation and scheduling protocols for wireless control systems that can both meet reliability \emph{and} latency requirements of the control system.
%
%
%In the wireless communications research and industry, many scheduling techniques have been proposed to provide reliability, or quality of service (QoS), to users across the network in the form of throughput, fairness and/or latency~\cite{cao2001scheduling,wongthavarawat2003packet,fattah2002overview,yaacoub2012survey}. For instance, round-robin scheduling is a common approach due to its simplicity inherent fairness. Algorithms such as proportional fair (PF)~\cite{kim2005proportional}, max-SNR~\cite{knopp1995information} and their variations have been designed for maximizing the overall system throughput and/or fairness. However, these algorithms can perform very poorly with regards to ensuring that all of the admitted users' packet delivery rates and latency targets will be met. For time-sensitive applications such as industrial control and automation, the experiences of individual users/devices are highly dependent on the ability of the network to deliver their packets within a latency bound. If this objective is not met, the high overall system throughput or throughput-fairness means little to the end users. On the other hand, the generic delay-aware schedulers such as EDF~\cite{wu2014analysis} and WFQ~\cite{lu1999fair} do not properly adapt to wireless channel conditions and may make poor radio resource allocation decisions. These algorithms have also been modified to adapt to wireless channel conditions, such as in M-LWDF \cite{andrews2001providing}. However, all such wireless scheduling algorithms are unable to leverage the information on the control system dynamics and thus, perform inefficiently when deployed in time-sensitive wireless control systems.
%
%In the context of wireless control systems, there was been a range or works that incorporates control system information in the networking and communication policies.
%%
%The mechanisms usually examined are either static or dynamic. Typical examples of the first type are periodically protocols where the wireless devices transmit in a predefined repeating order, e.g., round-robin. Control system stability under such protocols can be analyzed -- see, e.g.,~\cite{Hespanha_survey, Schenato_foundations, Donkers_switched, Branicky_stability}. Periodic sequences leading to stability~\cite{Hristu_shared_feedback}, controllability and observability~\cite{Hristu_communication_control}, or optimizing control objectives~\cite{LeNy_resource_LQR,Meier_measurement_control, Scheduling_control_combinatorics} have been proposed.
%%
%The second type of schedulers, the dynamic ones, do not rely on a predefined sequence but decide access to the communication medium dynamically at each step. %, for example by dynamically assigning priorities to the competing tasks.
%Initial approaches abstract control performance requirements in the time/frequency domain, e.g., how often a task needs resource access, employing algorithms from real-time scheduling theory~\cite{Branicky_RM, Liu_Real_time_systems}. More recently scheduling approaches often depend on the current control system states, i.e., informally the subsystem with the largest state discrepancy is scheduled to communicate -- see, e.g.,~\cite{Donkers_switched, Walsh_stability, Hristu_Kumar_interrupt_based, Egerstedt_queue, Hirche_Scheduling_Price, Cervin_event_scheduling, mamduhi2014event,shi2011optimal,han2017optimal}.
%%
%Alternatively scheduling can take into account current wireless channel conditions opportunistically to meet target control system reliability requirements~\cite{GatsisEtal15}. None of these approaches, however, are explicitly designed to minimize communications latency or achieve strong performance with specific low-latency requirements.
%
%
%
%In this paper, we develop a control-aware, high reliability, low-latency IEEE 802.11ax WiFi protocol~\cite{bellalta2016ieee} that is designed to reduce the total transmission time at each uplink cycle in the control loop. This is done through the mathematical formulation of the control system design goal in the form of a Lyapunov function that ensures stability of the control system. This formulation naturally induces a bound on the packet delivery rate each control system needs to achieve to meet the control-based goal. Such packet delivery rates depend upon current control and channel states and thus dynamically change over the course of the system life time. This can be viewed as an opportunistic protocol with respect to both the current control states \emph{and} channel conditions. Furthermore, because these control-based success rate requirements may be significantly lower than traditional, high reliability communication demands, the proposed method is better suited to find scheduling configurations that can moreover meet the strict latency requirements imposed by the physical system.
%
%The paper is organized as follows. We formulate the wireless control system in which state information is communicated to the control over a wireless channel. Due to the potential for random packet drops, this is modeled as a switched dynamical system (Section \ref{sec_problem_formulation}). A Lyapunov function is used to evaluate the stability of the control state, and the uncertainty in this measurement grows the more consecutive packets are lost for a particular system.  We then discuss the scheduling parameters of the IEEE 802.11ax communication model (Section \ref{sec_comm_model}).
%%, which allows for centralized scheduling of transmissions through the use of varying bandwidth resource units, time-slotted PPDUs, and date rate selection via MCS selection (Section \ref{sec_comm_model}). 
%From there, we derive a mathematical formulation of the optimal scheduling problem (Section \ref{sec_optimal}). This can be formulation by minimizing a control cost with an explicitly latency constraint (Section \ref{sec_optimal_a}) or minimizing transmission time with an explicit control performance constraint (Section \ref{sec_optimal_b}).
%
%Using the control-constrained optimal scheduling formulation, we develop the control-aware low latency scheduling (CALLS) method (Section \ref{sec_calls}). The CALLS method uses current control states and channel conditions to derive dynamic packet success rates for each user (Section \ref{sec_psr}). In this way, control systems that are closest to instability will be given priority in the scheduling so that they may close their control loops. The scheduling procedure consists of a random user selection procedure to reduce the number of required PPDUs that incur significant overhead  (Section \ref{sec_rss}), followed by an assignment-method based scheduling of selected users to minimize total transmission time (Section \ref{sec_assignment}). The performance of the CALLS method is analyzed in a series of simulation experiments in which its performance is compared against a control-agnostic procedure (Section \ref{sec_simulation}). We demonstrate in numerous low-latency control systems that the control-aware method can support more users than the alternative and achieve more robust overall performance.

>>>>

%The reliability and robustness of wireless communication systems has become a key component in the design and implementation of large scale systems in the Internet-of-Things (IoT). For these systems to operate successfully, it is important that the data collected by sensors can be communicated throughout the IoT network. While the most reliable form of data communication is through wired connections, the scale and mobility of modern IoT settings has rendered the cost of installing and maintaining a wired network a significant challenge~\cite{wollschlaeger2017future}. Consequently, there is great interest in the design of wireless control systems that can achieve reliable performance~\cite{varghese2014wireless, li2017review}. Low-latency wireless transmissions are however a necessary feature in many wireless IoT systems, particularly those in industrial control~\cite{varghese2014wireless,popovski2018wireless, bennis2018ultra}.  The primary challenge is then the inherent trade-off that occurs between reliability and latency, as it is difficult to maintain high reliability while using higher data rates due to the stochasticity of the wireless channel. It is necessary to design resource allocation and scheduling protocols for wireless control systems that can both meet reliability \emph{and} latency requirements of the control system.

%In the wireless communications research and industry, many radio resource allocation schemes in the form of wireless scheduling techniques have been proposed to provide reliability, or quality of service (QoS), to users across the network in the form of throughput, fairness and/or latency~\cite{cao2001scheduling,wongthavarawat2003packet,fattah2002overview,yaacoub2012survey}. For instance, round-robin scheduling is a common approach due to its simplicity inherent fairness. Algorithms such as proportional fair (PF)~\cite{kim2005proportional}, max-SNR~\cite{knopp1995information} and their variations have been designed for maximizing the overall system throughput and/or fairness. For time-sensitive applications such as industrial control and automation, the experiences of individual users/devices are highly dependent on the ability of the network to deliver their packets within a latency bound. Generic delay-aware schedulers such as EDF~\cite{wu2014analysis} and WFQ~\cite{lu1999fair} do not adapt to wireless channel conditions, while channel state information is considered in M-LWDF \cite{andrews2001providing}. However, all such methods are unable to leverage the information on the control system dynamics and thus can make unsuitable radio resource allocation decisions when deployed in time-sensitive wireless control systems.

\red{The reliability and robustness of wireless communication systems has become a key component in the design and implementation of large scale systems in the Internet-of-Things (IoT). For these systems to operate successfully, it is important that the data collected by sensors can be communicated throughout the IoT network. While the most reliable form of data communication is through wired connections, the scale and mobility of modern IoT settings has rendered the cost of installing and maintaining a wired network a significant challenge~\cite{zand2012wireless,wollschlaeger2017future}. Consequently, there is great interest in the design of wireless control systems that can achieve reliable performance~\cite{varghese2014wireless, li2017review}. Low-latency wireless transmissions are however a necessary feature in many wireless IoT systems, particularly those in industrial control~\cite{zand2012wireless,varghese2014wireless,weiner2014design,popovski2018wireless, bennis2018ultra}.  The primary challenge in so-called ultra reliable low latency communications (URLLC) is then the inherent trade-off that occurs between reliability and latency, as it is difficult to maintain high reliability while using higher data rates due to the stochasticity of the wireless channel. It is necessary to design resource allocation and scheduling protocols for wireless control systems that can both meet reliability \emph{and} latency requirements of the control system.

Many radio resource allocation schemes in the form of wireless scheduling techniques have been proposed to provide reliability, or quality of service (QoS), to users across the network in the form of throughput, fairness and/or latency~\cite{yaacoub2012survey}. Algorithms such as proportional fair (PF)~\cite{kim2005proportional}, max-SNR~\cite{knopp1995information} and their variations have been designed for maximizing the overall system throughput and/or fairness. For URLLC or time-sensitive applications such as industrial control and automation, the experiences of individual users/devices are highly dependent on the ability of the network to deliver their packets within a latency bound. Generic delay-aware schedulers such as EDF~\cite{wu2014analysis} and WFQ~\cite{lu1999fair} do not adapt to wireless channel conditions, which are considered in M-LWDF \cite{andrews2001providing}. Within the URLLC framework, low-latency communications have been enabled through achieving various forms of diversity \cite{swamy2015cooperative,popovski2018wireless,nielsen2018ultra}.   However, all such methods are unable to leverage the information on the control system dynamics and thus can make unsuitable radio resource allocation decisions when deployed in time-sensitive wireless control systems.}

In the context of wireless control systems, there have been a range of works that incorporates control system information in the networking and communication policies.
%
The mechanisms usually examined are either static or dynamic. Typical examples of the former type are periodically protocols where the wireless devices transmit in a predefined repeating order, e.g., round-robin. Control system stability under such protocols can be analyzed -- see, e.g.,~\cite{Hespanha_survey, Schenato_foundations, Donkers_switched, Branicky_stability}. Periodic sequences leading to stability~\cite{Hristu_shared_feedback}, controllability and observability~\cite{Hristu_communication_control}, or optimizing control objectives~\cite{LeNy_resource_LQR,Meier_measurement_control, Scheduling_control_combinatorics} have been proposed.
%
Dynamic schedulers do not rely on a predefined sequence but decide access to the communication medium dynamically at each step. %, for example by dynamically assigning priorities to the competing tasks.
Initial approaches abstract control performance requirements in the time/frequency domain, e.g., how often a task needs resource access, employing algorithms from real-time scheduling theory~\cite{Branicky_RM, Liu_Real_time_systems}. More recent scheduling approaches often depend on the current control system states, i.e., informally the subsystem with the largest state discrepancy is scheduled to communicate -- see, e.g.,~\cite{Donkers_switched,  Cervin_event_scheduling, mamduhi2014event,shi2011optimal,han2017optimal}.
%
Alternatively scheduling can take into account current wireless channel conditions opportunistically to meet target control system reliability requirements~\cite{GatsisEtal15}. \red{None of these approaches, however, are explicitly designed to achieve strong performance it low-latency scenarios, which are critical in modern industrial control and factory settings \cite{brahmi2015deployment,yilmaz2015analysis}.}

In this paper, we develop a control-aware, high reliability, low-latency IEEE 802.11ax WiFi protocol~\cite{bellalta2016ieee} that is designed to reduce the total transmission time at each uplink cycle in the control loop. This is done through the mathematical formulation of the control system design goal in the form of a Lyapunov function that ensures stability of the control system. This formulation naturally induces a bound on the packet delivery rate each control system needs to achieve to meet the control-based goal. Such packet delivery rates depend upon current control and channel states and thus dynamically change over the course of the system life time. This can be viewed as an opportunistic protocol with respect to both the current control states \emph{and} channel conditions. Furthermore, because these control-based success rate requirements may be significantly lower than traditional, high reliability communication demands, the proposed method is better suited to find scheduling configurations that can moreover meet the strict latency requirements imposed by the physical system. \red{In contrast to more standard URLLC approaches of fixing high reliability targets, our approach can be seen as \emph{adaptive} reliability with respect to control state and dynamics. }

@@

Using this formulation, we develop the control-aware low latency scheduling (CALLS) method (Section \ref{sec_calls}). The CALLS method uses current control states and channel conditions to derive dynamic packet success rates for each user (Section \ref{sec_psr}). In this way, control systems that are closest to instability will be given priority in the scheduling so that they may close their control loops. The scheduling procedure consists of a random user selection procedure to reduce the number of required PPDUs that incur significant overhead  (Section \ref{sec_rss}), followed by an assignment-method based scheduling of selected users to minimize total transmission time (Section \ref{sec_assignment}). The performance of the CALLS method is analyzed in a series of simulation experiments in which its performance is compared against a control-agnostic procedure (Section \ref{sec_simulation}). \red{We demonstrate in numerous control systems that the control-aware, adaptive reliability approach can support more users than the alternative and achieve more robust overall performance.}
